A finite-state morphological transducer for Kyrgyz

### 2012-03-13 paper/paper.tex
update

\subsection{Morphological and orthographic words}
Normal tokenisation often takes white space to be the delimiter between ``words''.  For analysis and generation, however, there are two exceptions to this.  Some morphophonological processes work across the ``whitespace'' boundary, and some clitics which are written next to the previous word without a whitespace can be considered syntactically separate units (words) and follow standard morphophonologial processes.

Kyrgyz, like other turkic languages (Tatar, Chuvash, Kazakh), has morphophonological processes that work across the ``whitespace'' boundary, but unlike many other Turkic languages, they are not represented in the orthography.

Some clitics in Kyrgyz which can be considered separate words but are written together with the previous word include the question word `бы'; the focus particle `чы'; copula suffixes `мин', etc.; and the progressive auxiliary `жат' with certain verbs.
%FIXME: why copular suffixes?

One result of making transducer targetted at machine translation is that we want to be able to both analyse, and generate the same forms. .... %FIXME

>>>>

\subsection{Morphological and orthographic words}
A typical tokenisation strategy is to take white space to be the delimiter between `words'.  For analysis and generation, however, there are exceptions to this.  Some morphophonological processes work across the `whitespace' boundary, and some clitics which are written next to the previous word without a whitespace can be considered syntactically separate units (words) and follow standard morphophonologial processes.

The first exception does not apply to Kyrgyz, as unlike other Turkic languages (Tatar, Chuvash, Kazakh), although it has morphophonological processes that work across the `whitespace' boundary, these are not represented in the orthography.

Regarding the second exception, some clitics in Kyrgyz can be considered separate words, but are written together with the previous word. For example, the question word `бы'; the focus clitic `чы'; copula suffixes `мин', etc.; and the progressive auxiliary `жат' with certain verbs.

The analyser is designed to be used with the longest-match left-to-right (LRLM) tokenisation strategy, as described in \cite{garrido-alenda02}. Thus when more than one `word' needs to be output, they are joined with the \texttt{+} symbol, which is reserved for joining two `words' in one analysis, e.g. {\em өнүктүрөбүзбү} `TRANSLATION' is analysed as 
{\tt өнүк<v><tv><caus><aor><p1><pl>+бы<mod_qst>}

% garrido-alenda02

%FIXME: why copular suffixes?

%One result of making transducer targetted at machine translation is that we want to be able to both analyse, and generate the same forms. 
%.... %FIXME
